A Formal Model of Checked C

### 2022-07-19 sandp23/introduction.tex
Updates from Overleaf

\section{Introduction}
Vulnerabilities due to memory corruption, especially spatial memory corruptions, are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe} despite a large body of work that tries
to prevent them~\cite{song2019sanitizing}.
Vulnerabilities due to memory corruption, especially spatial memory corruptions, are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe} despite a large body of work that tries
to prevent them~\cite{song2019sanitizing}.
One way to handle this is to convert these programs into safe languages such as Java. However, it requires considerable development effort. Furthermore, the performance overhead prohibits them from being used in performance-critical settings such as Operating Systems.

With the increased support for high-performant safe system languages such as Rust and Go, we can have tools that convert C to a memory-safe language, e.g., Rust, a promising memory-safe systems
language~\cite{rustlang}. Unfortunately, Rust is too different from C to constitute a practical target~\cite{zeng2019identifying}, at least for now: the (best of breed) \texttt{c2rust} tool~\cite{c2rust,c2rusttalk} transliterates C to \emph{unsafe, non-idiomatic} Rust, and recent
work~\cite{emre2021c2rust} is able to take only small steps to close the safety gap and still requires considerable work from developers.

There are other retrofitting techniques such as CCured~\cite{necula2005ccured},
Softbound~\cite{nagarakatte2009softbound}, Low Fat
pointers~\cite{duck2016heap}, and Address Sanitizer
(ASAN)~\cite{serebryany2012addresssanitizer} aim to enforce spatial safety
automatically~\emph{without any developer effort}, by analyzing a C program and compiling it to include
run-time safety checks. Unfortunately, the resulting run-time overhead is too high
for deployment (between 60\%-200\%), especially in low-powered IoT devices where the usage of C-based system software (e.g., FreeRTOS) is still rampant.
~\fig{fig:memprotections} shows the summary of these existing approaches.

%\input{figures/existingmem}

\emph{Checked C}~\cite{tarditi2018checked} as a
practical target to which to convert legacy/active C code. 
Checked C
extends C with \emph{checked pointer
  types} which are restricted by the compiler to spatially safe uses (temporal safety is underway~\cite{checkedc:temporal}).
Such pointers have one of three possible types, \ptrT{T}, \arrT{T},
or~\ntarrT{T} (\ptr, \arr, and \ntarr{} for short), representing a
pointer to a single element, array of elements, or null-terminated
array of elements of type $T$, respectively. The latter two have an
associated \emph{bounds annotation}; e.g., a declaration \arrT{int}
\code{p} : \code{count(n)} says that \code{p} is a pointer to an \code{int} array
whose size is \code{n}.
%
Checked C's Clang/LLVM-based compiler represents checked
pointers as system-level memory words, i.e., without
``fattening'' metadata, ensuring backward compatibility.
The compiler uses these bounds annotations to add dynamic checks prior to
checked pointer accesses, to prevent spatial safety violations.
These run-time checks can often be proved redundant and removed by LLVM, yielding good performance.
%, and leverages LLVM to optimize away run-time
%checks, yielding good performance.
%
Tarditi et al.~\cite{tarditi2018checked} reported average
run-time overheads of 8.6\% on a small benchmark suite, and
\cite{duanrefactoring} found essentially no overhead when running
Checked C-converted portions of the FreeBSD kernel.

However, converting existing C programs to Checked C requires considerable effort~\cite{duanrefactoring}. 
Recently, Machiry et al. developed 3C~\cite{machiry2022c} that tries to automatically convert a program to Checked C by adding appropriate pointer annotations.
However, as described in 3C, complete automated conversion is infeasible and consequently requires the developer to convert some code regions manually.
Thanks to the backward compatibility of Checked C, the partially annotated program still enjoys the benefits of spatial memory safety on those regions using only Checked pointers (i.e., checked or safe regions).
But, the unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}.
%
Although the blameless proof exists~\cite{ruef2019achieving}, it does not state that spatial safety violations cannot happen in Checked regions but rather states that Checked regions~\emph{cannot be blamed for any spatial safety violations}.
For instance, a checked function expects a pointer to a buffer of five elements, but an unchecked region can call this function with a pointer to a buffer of three elements. This results in a spatial safety violation in the Checked region, but of course, the blame or the root cause is in the unchecked region.
Furthermore, since checked and unchecked regions execute in the same address space, spatial memory corruptions in unchecked regions can take down the complete program despite having checked regions.
We need an isolation mechanism to ensure that code executed as part of unchecked regions does not violate the safety guarantees in checked regions.
Such an isolation mechanism should have low/no overhead and be backward compatible (i.e., pointers should have no additional metadata), preserving the advantages of Checked C.

>>>>

\section{Introduction}
Vulnerabilities due to memory corruption, especially spatial memory corruptions, are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe} despite a large body of work that tries
to prevent them~\cite{song2019sanitizing}.
Converting these existing programs into safe languages, such as Java, requires considerable effort and is not (yet) feasible for performance-critical environments such as Embedded Operating Systems.
Although high-performant safe system languages such as Rust~\cite{rustlang} and Go exist, these a too different to constitute a practical target~\cite{zeng2019identifying}, and rewriting in these languages requires considerable effort.
Existing automated conversion tools~\cite{c2rust,c2rusttalk, emre2021c2rust} are able to take only small steps (i.e., to unsafe, non-idiomatic Rust) and still~\emph{require considerable work from developers}.
There are other retrofitting techniques such as CCured~\cite{necula2005ccured},
Softbound~\cite{nagarakatte2009softbound}, Low Fat
pointers~\cite{duck2016heap}, and Address Sanitizer
(ASAN)~\cite{serebryany2012addresssanitizer} aim to enforce spatial safety
automatically~\emph{without any developer effort}, by analyzing a C program and compiling it to include
run-time safety checks. Unfortunately, the resulting~\emph{run-time overhead is too high
for deployment (between 60\%-200\%)}, especially in low-powered IoT devices where the usage of C-based system software (e.g., FreeRTOS) is still rampant.

\noindent
\textbf{Checked C.} Tarditi et al. developed~\emph{Checked C}, a safe dialect of C that prevents spatial memory issues (temporal safety is underway~\cite{checkedc:temporal}) with no overhead.
Checked C extends C with \emph{checked pointer
  types} which are restricted by the compiler to spatially safe uses (temporal safety is underway~\cite{checkedc:temporal}).
Such pointers have one of three possible types, \ptrT{T}, \arrT{T},
or~\ntarrT{T} (\ptr, \arr, and \ntarr{} for short), representing a
pointer to a single element, array of elements, or null-terminated
array of elements of type $T$, respectively. The latter two have an
associated \emph{bounds annotation}; e.g., a declaration \arrT{int}
\code{p} : \code{count(n)} says that \code{p} is a pointer to an \code{int} array whose size is \code{n}.
The compiler uses these bounds annotations to add dynamic checks prior to checked pointer accesses, to prevent spatial safety violations.
These run-time checks can often be proved redundant and removed by LLVM, yielding good performance.
Specifically, Tarditi et al.~\cite{tarditi2018checked} reported average
run-time overheads of 8.6\% on a small benchmark suite, and
Duan et al.~\cite{duanrefactoring} found essentially~\emph{no overhead} when running
Checked C-converted portions of the FreeBSD kernel.
Furthermore, Checked C is backward compatible as the compiler  represents checked pointers as system-level memory words, i.e., without
``fattening'' metadata, ensuring backward compatibility. 
The backward compatibility also enables incremental or partial conversion where Checked annotations can be provided to a few pointers.
Such partially annotated programs still enjoy the benefits of spatial memory safety on those regions using only Checked pointers (i.e., checked or safe regions).
~\fig{fig:memprotections} shows the summary of these existing approaches along with Checked C.
We argue that~\emph{converting to Checked C provides a reasonable and practical approach to adding safety to existing system software written in C}.

\input{figures/existingmem}

\noindent
\textbf{Converting C to Checked C.} The safety guarantees of Checked C come with certain restrictions. For instance, as shown below, Checked C programs cannot use address-taken variables in a bounds expression as the bounds relations may not hold because of possible modifications through pointers.
\begin{minted}[mathescape, escapeinside=||, fontsize=\footnotesize}{c}
...
array_ptr<int> p : count (n) = NULL;
|\textcolor{red}{\faTimes}|..,&n,.
\end{minted}
Consequently, converting existing C programs to Checked C might require refactoring, e.g., modifying the above program to not use~\inlinecode{&n} expression, which might require considerable effort~\cite{duanrefactoring} depending on the program's complexity. 
Recently, Machiry et al. developed~\threec~\cite{machiry2022c} that tries to automatically convert a program to Checked C by adding appropriate pointer annotations.
However, as described in \threec, complete automated conversion is infeasible and requires the developer to convert some code regions manually.
Although, the backward compatibility of Checked C helps a partially annotated program to enjoy spatial memory safety on those regions using only Checked pointers (i.e., checked or safe regions).
But, the unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}.
Although the blameless proof exists~\cite{ruef2019achieving}, it does not state that spatial safety violations cannot happen in Checked regions but rather states that Checked regions~\emph{cannot be blamed for any spatial safety violations}.
Consider the following example:
\begin{minted}[mathescape, linenos, escapeinside=||, fontsize=\footnotesize{}]{c}
// Checked code
int func(array_ptr<char> p : count(5)) {
|\textcolor{red}{\faChainBroken}|..p[4]..
}

// unchecked code
...
str = "he";
...
|\textcolor{red}{\faBug}|assume_bounds_cast<char>(str, 5); 
...
char ptr[16];
...
len <- derived from user input
...
|\textcolor{red}{\faBug}| memcpy(ptr, buff, len);
\end{minted}
Here, the checked function~\inlinecode{func} expected a pointer to a buffer of five elements, but unchecked code violated it and invoked the function with a buffer of 2 elements.
This results in a spatial safety violation (\textcolor{red}{\faChainBroken}) in the Checked region, but of course, the blame or the root cause is in the unchecked region (\textcolor{red}{\faBug}).
Furthermore, since checked and unchecked regions execute in the same address space, spatial memory corruptions in unchecked regions (line ...) can take down the complete program despite having checked regions.
We need an isolation mechanism to ensure that code executed as part of unchecked regions does not violate the safety guarantees in checked regions.
However, converting existing C programs to Checked C requires considerable effort~\cite{duanrefactoring}. 
Recently, Machiry et al. developed 3C~\cite{machiry2022c} that tries to automatically convert a program to Checked C by adding appropriate pointer annotations.
However, as described in 3C, complete automated conversion is infeasible and consequently requires the developer to convert some code regions manually.
Thanks to the backward compatibility of Checked C, the partially annotated program still enjoys the benefits of spatial memory safety on those regions using only Checked pointers (i.e., checked or safe regions).
But, the unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}.
%
Although the blameless proof exists~\cite{ruef2019achieving}, it does not state that spatial safety violations cannot happen in Checked regions but rather states that Checked regions~\emph{cannot be blamed for any spatial safety violations}.
For instance, a checked function expects a pointer to a buffer of five elements, but an unchecked region can call this function with a pointer to a buffer of three elements. This results in a spatial safety violation in the Checked region, but of course, the blame or the root cause is in the unchecked region.
Furthermore, since checked and unchecked regions execute in the same address space, spatial memory corruptions in unchecked regions can take down the complete program despite having checked regions.
We need an isolation mechanism to ensure that code executed as part of unchecked regions does not violate the safety guarantees in checked regions.
Such an isolation mechanism should have low/no overhead and be backward compatible (i.e., pointers should have no additional metadata), preserving the advantages of Checked C.
